The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum

### 2014-03-20 cordillera-climate.tex
Typesetting: Fix Fig., Table and Sect. references

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}. Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively. Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}. On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.

@@

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs. \ref{fig:ivolarea}-\ref{fig:cool05}).

@@

As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}). In this section we use the WorldClim climatology, which is derived from observations, as a~reference with which to compare other climate forcing data on land. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}). Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).

@@

Most likely, this is the signature of an orographic precipitation effect. As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

@@

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR). These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

% fig:topo
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-topo}
	\end{center}
	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}. Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a~reference for temperature lapse-rate corrections. Fig.~\ref{fig:topo}-\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:topo}
\end{figure*}

@@

% fig:prec
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-prec}
	\end{center}
	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies. Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}
	\label{fig:prec}
\end{figure*}

@@

% fig:biatm
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatm}
	\end{center}
	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with figure~\ref{fig:cool05}}
	\label{fig:biatm}
\end{figure*}

% fig:biatmbars
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatmbars}
	\end{center}
	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation. Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
	\label{fig:biatmbars}
\end{figure}

>>>>

In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}. Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table~\ref{tab:reanalyses}).

The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figs.~\ref{fig:temp} and~\ref{fig:prec}. Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively. Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}. On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.

@@

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).

@@

As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}). In this section we use the WorldClim climatology, which is derived from observations, as a~reference with which to compare other climate forcing data on land. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}). Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

Most likely, this is the signature of an orographic precipitation effect. As described in Sect.~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

@@

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR). These qualitative “best” runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

% fig:topo
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-topo}
	\end{center}
	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}. Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a~reference for temperature lapse-rate corrections. Figs.~\ref{fig:topo}--\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
	\label{fig:topo}
\end{figure*}

@@

% fig:prec
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-prec}
	\end{center}
	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies. Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (Sect.~\ref{sec:climate}).}
	\label{fig:prec}
\end{figure*}

@@

% fig:biatm
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatm}
	\end{center}
	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
	\label{fig:biatm}
\end{figure*}

% fig:biatmbars
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatmbars}
	\end{center}
	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation. Corresponding ice sheet geometries are presented in Figs.~\ref{fig:cool05} and~\ref{fig:biatm}.}
	\label{fig:biatmbars}
\end{figure}
